---
title: "R Notebook"
output:
  html_document: default
  html_notebook: default
---

# Packages to be installed

```{r echo=FALSE}
if (!require(tidyverse)){install.packages("tidyverse",dependencies = TRUE)} 
library(tidyverse)

if (!require(readr)){install.packages("readr",dependencies = TRUE)} 
library(readr)

if (!require(lubridate)){install.packages("lubridate",dependencies = TRUE)} 
library(lubridate)

if (!require(ggplot2)){install.packages("ggplot2",dependencies = TRUE)} 
library(ggplot2)

if (!require(randomForest)){install.packages("randomForest",dependencies = TRUE)} 
library(randomForest)

if (!require(party)){install.packages("party",dependencies = TRUE)} 
library(party)

if (!require(ipred)){install.packages("ipred",dependencies = TRUE)} 
library(ipred)

if (!require(forecast)){install.packages("forecast",dependencies = TRUE)} 
library(forecast)

```

## Import the data from source location:

```{r}
bookings <- read_csv("C:/Users/sayooj/Downloads/bookings.csv")
#View(bookings)
```

## Tidy the data

```{r}

bookings <- bookings %>% drop_na()

```




# Feature Engineering

### Parse Out Day of Week

```{r}

bookings$DayOfDeparture  = wday(mdy_hm(bookings$DepartureDate,locale = "English"),label = TRUE)
```

### Normalize Fare Per Person
```{r}
bookings$FarePP = bookings$Fare/bookings$PassengerCount
```

### Convert to FACTOR

```{r}
bookings$class = as.factor(bookings$InvClass)
bookings$orig = as.factor(bookings$Origin)
bookings$desti = as.factor(bookings$Destination)
```

### Puzzling FE

```{r}
varNames <- names(bookings)
# Exclude ID or Response variable
varNames <- varNames[varNames %in% c("FarePP","ID","FareBreakFareBasis","BookingAgent",
                                      "TicketDate","AgencyNumber","Channel","DepartureDate","Fare",
                                      "InvClass","Origin","Destination","PassengerMiles")]

# add + sign between exploratory variables
varNames1 <- paste(varNames, collapse = "+")
varNames1

# Add response variable and convert to a formula object
rf.form <- as.formula(paste("FarePP", varNames1, sep = " ~ "))
rf.form

v1 <- c(bookings$class,bookings$orig,bookings$desti,bookings$PassengerCount,
          bookings$PassengerMiles,bookings$DayOfDeparture)

# is.finite(v1)
# which(is.infinite(v1))
# is.na(v1)
# which(is.na(v1))

```



### Time-series plot of Fare for every destination

```{r}
ggplot(na.omit(bookings),aes(mdy_hm(DepartureDate,locale = "English"),FarePP)) + 
  geom_point() + 
  facet_wrap(~ Destination) + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

### Time-series plot of Fare for every origin

```{r}
ggplot(na.omit(bookings),aes(mdy_hm(DepartureDate,locale = "English"),FarePP)) + 
  geom_point() + 
  facet_wrap(~ Origin) + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```


### Split Data into 3 - train, test and validation

```{r}
newdata = sample(seq(1, 3), size = nrow(bookings), replace = TRUE, prob = c(.4, .3, .3))
train = bookings[newdata == 1,]
validation = bookings[newdata == 2,]
test = bookings[newdata == 3,]
```

# Modeling

```{r}
rmse <- function(error)
{
    sqrt(mean(error^2))
}
```


## Ensemble Methods

### Bagged Trees
```{r}
baggedTree.time0 = proc.time()

baggedTree = bagging(FarePP~orig+desti+PassengerMiles+DayOfDeparture+class,data = bookings)

baggedTree.time = proc.time()-baggedTree.time0

bookings.pred.bagging = predict(baggedTree,bookings)

# cbind(bookings$FarePP,bookings.pred.bagging)
```




## Forecasting Fares

### Data Prep

```{r}
bookings %>% group_by(orig,desti) %>% tally(sort = TRUE) #indicates DATASET has DTW-LAS as highest number of flights


df_LAS = bookings %>% dplyr::filter(orig == 'DTW' & desti == 'LAS') %>% 
              select(DepartureDate,FarePP) %>% 
              filter(FarePP > 0) %>% 
              group_by(DepartureDate) %>% 
              summarise(FarePP = mean(FarePP))

df_LAS$DepartureDate = mdy_hm(df_LAS$DepartureDate,locale = "English")

tsdf_LAS = ts(df_LAS$FarePP,start = 2005, frequency = 365.25)

tsdf_LAS
```


### Exponential Smoothing Method
```{r}
m_ets = ets(tsdf_LAS)
f_ets = forecast(m_ets, h=15) # forecast 15 months into the future
plot(f_ets)
```


### Auto ARIMA
```{r}
m_aa = auto.arima(tsdf_LAS)
f_aa = forecast(m_aa, h=6) # forecast 24 months into the future
plot(f_aa)
```


### Forecasting Model Comparison

```{r}
barplot(c(ETS=m_ets$aic, ARIMA=m_aa$aic),
    col="light blue",
    ylab="AIC")
```





